---
title: "Introduction to Public Health Data Science Using R"
author: "LB"
date: "2021/03/06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the Tidyverse
```{r tv}
install.packages("tidyverse", dependencies = TRUE)
```

It's a big one but this one set of packages is all we will need for the entirety of this workshop! :)

### {dplyr}

dplyr is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges.

#### 1. `filter()`

`filter()` does exactly what it sounds like. It filters your data down to only the *observations* you want to **keep**.
```{r filter, message=FALSE, warning=FALSE}
#load the data into a global enviroment
data("mpg")
#load the dplyr package
library("dplyr")
#`?mpg`

#filter(dataObject, variableConditions) 
filter(mpg, hwy >= 30)
```

Remember the `<-` operator. It would be a lot more helpful to store our new data so that we can use it later.

```{r filter2}
goodMPG <- filter(mpg, hwy >= 20)
```

Now *YOU* try it. Explore the `mpg` dataset and filter on a variable.. or two if you want to get fancy ;)

```{r filter_you, eval=FALSE, message=FALSE, warning=FALSE}
myData <- filter(mpg, [ADD YOUR CONDITION(S) HERE])
#if you do want to filter on more than one condition, just separate by a comma. Commas are treated as AND.
```

#### 2. `select()`

`select()` keeps all your observation but subsets the data down to only the *variables* you want to keep.

Remeber the `%>%` operator? Let's say I want to keep the only the cars that get more than 20 MPG but I also don't want to create a new dataframe when I reduce it to only the variables I need? This is where the pipe comes in. It allows me to pass the results of one step into another function! It makes your code look more *elegant* and saves on *memory*!

```{r select}
goodMPG <- filter(mpg, hwy >= 30) %>% 
  #feel free to replace these variable with your own set!
  select(hwy, displ, year, cyl, cty, class) #tip: type ?mpg in the console to see more information on the dataset
```

#### 3. `mutate()`

`mutate()` might be one of the most commonly used functions. This is the function that lets you create new variables!

Lets create the combined gas mileage. A simple Google search told me that "Combined fuel economy is a weighted average of City and Highway MPG values that is calculated by weighting the City value by 55% and the Highway value by 45%". 

```{r mutate}
goodMPG <- filter(mpg, hwy >= 30) %>% 
  select(class, displ, year, cyl, cty, hwy) %>% 
  #remember to give your variables meaningful names that are short and consice
  mutate(combined_mpg = round((.55*cty)+(.45*hwy)))
```

#### 4. `arrange()`

`arrange()` arranges :) It orders the rows of a data frame by the values of selected columns.

```{r arrange}
goodMPG <- filter(mpg, hwy >= 30) %>% 
  select(class, displ, year, cyl, cty, hwy) %>% 
  mutate(combined_mpg = round((.55*cty)+(.45*hwy))) %>% 
  #use the desc() to order in descending order, feel free to arrange in ascending!
  arrange(desc(combined_mpg))
```

#### 5. `summarise()`

`summarise()` is very fun function. It allows you to summarize your data based on some (or none) grouping variables. However, be mindful that summarize returns a different dataframe than the input. It will have one (or more) rows for each combination of grouping variables; if there are no grouping variables, the output will have a single row summarizing all observations in the input.

We will discuss how to implement grouping in a later step. 

```{r summarise}
# I want to save this as a separate object since I will likely need my tidy data later

averageMPG <- filter(mpg, hwy >= 30) %>% 
  select(class, displ, year, cyl, cty, hwy) %>% 
  mutate(combined_mpg = round((.55*cty)+(.45*hwy))) %>% 
  arrange(desc(combined_mpg)) %>% 
  #take the average combined mpg of all cars that get >= 30 MPG
  summarise(avg_combined = mean(combined_mpg))
```

*Q1:Run the above code chunk. What does this step return? Why?*

## Let's Plot!

One of the most powerful functions in the tidyverse is `ggplot2`. You can create beautiful plots with just one package! 
A few notes about `ggplot()`  
- `ggplot()` is designed to work with data frames as the data source, not individual vectors.  
- Internal syntax that is different from the tidyverse.  
- You can add as many layers to a plot. Here is a typical template:

```{r ggplot, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

Now let's try it.

```{r bar_plot, echo=FALSE}
#load the library
library(ggplot2)

#Geoms are added to your plot using `+` sign BUT you can pipe into ggplot.
## let's start with a simple bar plot looking at difference in cylinders
goodMPG %>% 
  ggplot(aes(x = class)) +
  geom_bar()

#now let's look at the difference between cyl and mpg
goodMPG %>% 
  ggplot(aes(x = class, y = combined_mpg)) +
  geom_col()
```

*Q2: What is happening in this second plot? Take a minute to think about why this plot is not what you would expect? (hint: try typing `?geom_col` and `?geom_bar` in the console to understand the differences between these two functions)*

____________________________
## Our Data! 

Now that we've got some basics of R and the Tidyverse covered. Let's actually work with some real health data and apply these principals as data scientists do!

### 1. Import
```{r import, message=FALSE, warning=FALSE}
library(readr)
#important to use RELATIVE PATHS not ABSOLUTE
fetal <- read_csv("./data/fetal_health.csv")

str(fetal)
```

*Q3: What do you think `str()` does? What can we say about this data? What are our variable classes? What is our N?*

### 2. Tidy

*Q4: Given what we know about the principals of tidy data, is our data tidy?*

### 3. Transform
Before we can even begin this step, what are we missing? A research question!
Yes, it is possible to perform non-hypothesis driven analyses but for the purposes of this workshop, it would be more helpful to have a hypothesis.  